Modal Abstractions for Virtualizing Memory Addresses

### 2024-11-13 programlogic.tex
more edits, adjusting presentation of some code figures, and removing some redundant text now has the experiments nearly fitting in 20 pages, but then there is related work after that still

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity we refer to the ghost map summarizing virtual-to-physical translations by 
\[\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)\]
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical
 $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in the virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\mathsf{va}}{\mathsf{paddr}}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
guaranteeing that ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant contains the physical resources that guarantee the hardware resolves the translation.

@@

Our logic is the first to address the existence of multiple address spaces, so requires
a way to look up
the ghost map of virtual-to-physical translations for an address space (the ghost name $\delta$ in Figure \ref{fig:virtualpointstosharing}).
% for each address space must be locateable from a representation of the specific address space:
% This means we also require a way to locate the map for the current address space from 
We support lookup by the current root page table pointer stored in \lstinline|cr3|.
\looseness=-1

@@

\paragraph{From A Single Address Space to Many}
In principle the resources from each individual address space should be collected into a single
shared invariant, and for each memory access,
a fragment of this corresponding to $\mathcal{I}\textsf{ASpace}(\Theta,m)$
for the current address space should be extracted from this global resource, used to prove correctness of the
memory access, and put back.
In this paper we focus on the middle section, explicity identifying resources for each
address space. The reason for this is that in practice, this global resource is also
the place where kernel-specific assumptions, such as guaranteeing that certain virtual address range
was mapped in all address spaces, would be enforced. This paper focuses on the
reasoning principles behind the virtual memory access, and we leave the use of this within a larger
kernel to future work. We do, however, deal with the existence of multiple address spaces in examples (Section \ref{sec:experiment}).

@@

@@

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
% Per the discussion in Section \ref{sec:issues}, w
As common for assembly-level verification, we define our logic using Hoare doubles:\footnote{This
omits some low-level Iris details (stuckness, observations) that play no meaningful
role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is Iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} an explicit parameter to the double, rather than simply using register assertions
directly 
makes it possible to ensure ownership of the \lstinline|cr3| register and its value is accounted for
while avoiding some technical headaches with trying to enforce that $\Phi$ itself contains that.
% solves a technical problem with ensuring that the page table root used to evaluate
% the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
% is feasible.
% \footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
% opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
% require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
% but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
%   difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
%   \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
%   significant proof burden.
% }

The rest of this section describes specifications of three key \textsf{AMD64} instructions 
in our logic. 
These rules and others (e.g., including accessing memory at an instruction-specified offset from a register
value, which is common in most ISAs)
can be found in our artifact.
Each rule in Figure \ref{fig:wpdamd}, 
is annotated with a root (i.e., \lstinline|cr3|) address value (\textsf{rtv}), 
under which the resources mentioned in the specification are valid.
In general, we use metavariables $\textsf{r}_s$ and $\textsf{r}_d$ to specify source and destination registers
for each instruction, and prefix various register value variables with \textsf{rv}.
We sometimes use $\textsf{r}_a$ to emphasize when a register is expected to hold an address used
for memory access, though the figure also uses typical assembler conventions of specifying
memory access operands by bracketing the register holding the memory address.
Standard for Hoare doubles, there is a frame resource $P$ in each rule for passing resources
not used by the first instruction in sequence through to subsequent instructions.
Our rules include tracking of each instruction's memory address to track \lstinline|rip| updates, which is critical
for control transfer instructions. Our development also includes handling of the \lstinline|rflags| register updates from arithmetic instructions.
Most rules are otherwise standard (e.g., \lstinline|mov| between register, etc.), with Figure \ref{fig:wpdamd} showing the rules
most unique to our development.
\looseness=-1

@@

\caption{Reasoning Rules for Selected \textsf{AMD64} Instructions under}

@@

\subsection{Soundness}
Our rules from Figure \ref{fig:wpdamd} are proven sound in Iris against an assembly-level hardware model
implementing a fragment of x86-64 including 64-bit address translation with 4-level page tables.
\todo[inline]{
Our rules for control transfers (\lstinline|jmp|, \lstinline|call|, and \lstinline|ret|) are currently
axiomatized (with completely standard specifications~\cite{ni2007contexts,Chlipala2013Bedrock}) because Iris's built-in machinery does not provide
convenient ways to discard the current continuation. We are working to adapt existing machinery~\cite{de2023type} to prove their soundness.
Our soundness proofs for all other instructions (including the instructions above, arithmetic, other heap-accessing operations, push, and pop)
are axiom-free.
}
% At the moment our proofs do rely on 13 small axioms of properties which should be provable, but
% are challenging to discharge due to some representation choices in our model.\footnote{See \lstinline|srx/x64/machine/current_axioms.v|}

>>>>

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity we refer to the ghost map summarizing virtual-to-physical translations by 
\centerline{$\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)$}\\
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical
 $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in the virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\mathsf{va}}{\mathsf{paddr}}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
guaranteeing that ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant contains the physical resources that guarantee the hardware resolves the translation.

@@

Our logic is the first to address the existence of multiple address spaces, so also requires
a way to look up
the ghost map of virtual-to-physical translations for an address space (the ghost name $\delta$ in Figure \ref{fig:virtualpointstosharing}).
% for each address space must be locateable from a representation of the specific address space:
% This means we also require a way to locate the map for the current address space from 
We support lookup by the current root page table pointer stored in \lstinline|cr3|.
\looseness=-1

@@

% \paragraph{From A Single Address Space to Many}
% In principle the resources from each individual address space should be collected into a single
% shared invariant, and for each memory access,
% a fragment of this corresponding to $\mathcal{I}\textsf{ASpace}(\Theta,m)$
% for the current address space should be extracted from this global resource, used to prove correctness of the
% memory access, and put back.
% In this paper we focus on the middle section, explicity identifying resources for each
% address space. The reason for this is that in practice, this global resource is also
% the place where kernel-specific assumptions, such as guaranteeing that certain virtual address range
% was mapped in all address spaces, would be enforced. This paper focuses on the
% reasoning principles behind the virtual memory access, and we leave the use of this within a larger
% kernel to future work. We do, however, deal with the existence of multiple address spaces in examples (Section \ref{sec:experiment}).

@@

In general, per-address-space invariants should be collected in a larger
VMM invariant, with individual address spaces' invariants pulled out as needed, such as when proving
soundness of an individual virtual memory access.
But such larger invariants would contain many kernel-specific properties that are orthogonal
to the fundamental reasoning principles that are the focus of this paper.
We leave such kernel-specific reasoning to future work, but our verification of task switching
(Section \ref{sec:experiment}) demonstrates support for managing multiple address spaces.
\looseness=-1

@@

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
% Per the discussion in Section \ref{sec:issues}, w
As common for assembly-level verification, we define our logic using Hoare doubles:%\footnote{This
% omits some low-level Iris details (stuckness, observations) that play no meaningful
% role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is Iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} an explicit parameter to the double, rather than simply using register assertions
directly 
makes it possible to ensure ownership of the \lstinline|cr3| register and its value is accounted for
while avoiding some technical headaches with trying to enforce that $\Phi$ itself contains that.
% solves a technical problem with ensuring that the page table root used to evaluate
% the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
% is feasible.
% \footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
% opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
% require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
% but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
%   difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
%   \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
%   significant proof burden.
% }

The rest of this section describes specifications of three key \textsf{AMD64} instructions 
in our logic. 
These rules and others (e.g., including accessing memory at an instruction-specified offset from a register
value, which is common in most ISAs)
can be found in our artifact.
% Each rule in Figure \ref{fig:wpdamd}, 
% is annotated with a root (i.e., \lstinline|cr3|) address value (\textsf{rtv}), 
% under which the resources mentioned in the specification are valid.
In general, we use metavariables $\textsf{r}_s$ and $\textsf{r}_d$ to specify source and destination registers
for each instruction, and prefix various register value variables with \textsf{rv}.
We sometimes use $\textsf{r}_a$ to emphasize when a register is expected to hold an address used
for memory access, though the figure also uses typical assembler conventions of specifying
memory access operands by bracketing the register holding the memory address.
Standard for Hoare doubles, there is a frame resource $P$ in each rule for passing resources
not used by the first instruction in sequence through to subsequent instructions.
Our rules include tracking of each instruction's memory address to track \lstinline|rip| updates, which is critical
for control transfer instructions. Our development also includes handling of the \lstinline|rflags| register updates from arithmetic instructions.
Most rules are otherwise standard (e.g., \lstinline|mov| between register, etc.), with Figure \ref{fig:wpdamd} showing the rules
most unique to our development.
\looseness=-1

@@

\caption{Proof rules for selected \textsf{AMD64} instructions}

@@

\subsection{Soundness}
Our rules from Figure \ref{fig:wpdamd} are proven sound in Iris against an assembly-level hardware model
implementing a fragment of x86-64 including 64-bit address translation with 4-level page tables.
% \todo[inline]{
Our rules for control transfers (\lstinline|jne|, \lstinline|call|, and \lstinline|ret|) are currently
axiomatized (with completely standard specifications~\cite{ni2007contexts,Chlipala2013Bedrock}) 
because Iris's built-in machinery does not provide
convenient ways to discard the current continuation; adaptation of others'
approaches~\cite{de2023type} is ongoing work.
Our soundness proofs for all other instructions (including, critically, all memory accesses)
are axiom-free.
% }
% At the moment our proofs do rely on 13 small axioms of properties which should be provable, but
% are challenging to discharge due to some representation choices in our model.\footnote{See \lstinline|srx/x64/machine/current_axioms.v|}
\looseness=-1
